The \proglang{R} Package \pkg{forestinventory}: Design-Based Global and Small Area Estimations for Multiphase Forest Inventories

### 2017-08-31 rnw/Threephase_Estimators and their Application.Rnw
update Andreas 29.08.2017(2)
update Andreas 31.08.2017

The internal model case again insures that the zero mean residual property over $F$, and the \textbf{point estimates} are thus calculated as

@@

Note that the \code{grisons} inventory dataset originally corresponds to a classical two-phase sample design, i.e. the full set of explanatory variables is available at all sample locations in the inventory area. To demonstrate the three-phase estimation procedure, we provide an artificial three-phase scenario by recoding the phase indicators in the data set (column \code{phase_id_3p}). The values for indicating the phases (\code{0,1,2}) follow the terminology used in this article ($s_0$, $s_1$, $s_2$).

We can use the \code{table()} function to see the sample sizes of all phases for the three-phase secenario:

@@

We see that in the three-phase example, we now have 40 terrestrial sample locations providing the timber volume per hectare (\code{tvol}). The large phase $s_0$ comprises $178 + 88 + 40 = 306$ sample locations, since $s_2 \subset s_1 \subset s_0$. We now assume that at sample locations $x \in s_0$, only the mean canopy height (\code{mean}) is available, whereas at sample locations $x \in s_1$ we have also have the explanatory variables \code{stddev}, \code{max} and \code{q75} at hand. We can now define the reduced and full regression model formulas that we will use in the \code{threephase()} function. Note that the models are nested.

@@

Note that compared to the \code{twophase()} function, we now have to specify the two regression models, i.e. the nested reduced (\code{formula.s0}) and full (\code{formula.s1}) regression model. In addition, we also have to specify the indication of the $s_1$ phase (\code{s1.id}) in the argument \code{phase_id}. The global three-phase estimation can thus be applied by

@@

The summary of a \code{threephase} function recalls both regression model formulas and also gives the coefficients of determination $R^2$ for both the reduced (\code{r.squared_reduced}) and the full (\code{r.squared_full}) regression model. The set of sample sizes have also been extented by that of the $s_0$ phase (\code{n0}). In order to compare the results to the respective two-phase estimation, we assume that we have the mean canopy height as the only explanatory variable at all 306 sample locations:

@@

% ---------------------------------------------------------------------------- %
\subsection{Small Area Estimators}

@@

The \textbf{small} (\textit{small}) and \textbf{pseudo small area estimator} (\textit{psmall}) follow the same principle as in the two-phase case, i.e. applying the globally derived reduced and full regression model to the small area means of the explanatory variables, and then correcting for a potential model bias by adding the residual mean of the full model, which are not zero in general, to the point estimate (equations \ref{eq:pe_3p_small} and \ref{eq:pe_3p_psmall}). Note that the difference between the mean $\hat{\bar{\pmb{Z}}}^{(0)}_{1,G}$ derived in the $s_1$ phase and the (more) exact mean $\hat{\bar{\pmb{Z}}}^{(0)}_{0,G}$ or $\bar{\pmb{Z}}^{(0)}_{0,G}$ derived in the $s_0$ phase is again considered as a residual term likewise in the global estimation (equation \ref{eq:reg3p_ex} and \ref{eq:reg3p_nex}).

@@

The \textbf{synthetic} (\textit{synth}) and \textbf{pseudo synthetic estimator} (\textit{psynth}) ca be applied if no terrestrial samples are available in the small area, i.e. $n_{2,G}=0$. Consequently, the residual correction and the residual variation term of the full model can no longer be applied an vanishes from the point (equation \ref{eq:pe_3p_synth} and \ref{eq:pe_3p_psynth}) and design-based variance (equation \ref{eq:var_3p_reg_synth} and \ref{eq:var_3p_reg_psynth}) formulas. The point estimates are again potentially biased since the zero mean residual property $\int_G R(x)dx=0$ of the full model can not be insured within the small area $G$. Also the variance will be small but to the cost of ignoring the model uncertainties. Note that there is again no external variance formula for the synthetic and pseudo synthetic estimation.

@@

The \textbf{extended synthetic} (\textit{extsynth}) and \textbf{extended pseudo synthetic estimator} (\textit{extpsynth}) insures that the residuals of the full model over both the entire inventory area $F$ and the small area $G$ are zero at the same time, i.e. $\int_F R(x)dx=\int_G R(x)dx=0$. This is again realized by extending the vector of explanatory variables by a binary categorical indicator variable $I_G(x)$ which takes the value \code{1} if the observation lies inside the small area $G$ and is otherwise set to \code{0}. The extended auxiliary vector thus is defined as $\pmb{\mathbb{Z}}^t(x)=(\pmb{\mathbb{Z}}^{(0)t}(x),\pmb{\mathbb{Z}}^{t}(x))$, where $\pmb{\mathbb{Z}}^{(0)t}(x)=(\pmb{Z}^{(1)t}(x), I_G^t(x))$ and $\pmb{\mathbb{Z}}^{t}(x)=\pmb{Z}^{t}(x)$. This practically means that we have to specify the binary indicator variable for all observations in the data frame that is passed to the package-functions.

@@

We will demonstrate the use of three-phase small area estimation in the package \pkg{forestinventory} by applying the \textbf{extended synthetic} and the \textbf{synthetic} estimator to the \code{grisons} dataset. The setup is thus exactly the same as in the example for global three-phase estimation (section \ref{sec:glob_est_3p}). However, this time will use the exact auxiliary mean of the mean canopy height variable (\code{mean}) and assume that we do not know the exact means of the remaining explanatory variables \code{stddev}, \code{max} and \code{q75}. We thus first define the true means for each small area likewise we did in the \code{twophase()} example (section \ref{sec:twophase_sae}):

@@

We see that the \code{threephase()} function returns the sample sizes in the entire inventory area as well as within each small area. The value \code{Inf} for \code{n0G} indicates that the explanatory variables at $s_0$ sample locations used in the reduced model were in our case derived exhaustively. If we compare the two results, we see that the synthetic estimation again yields a much lower variance than the extended synthetic estimation, but to the cost of ignoring the model uncertainties.

We can also analyse how the exhaustive derivation of \code{mean} performed compared to the case where \code{mean} is not non-exhaustively available but only at a very large $s_0$ phase with $n_{0,G}>>n_{1,G}$. To do this, we additionally compute the \textbf{extended pseudo synthetic} estimates. As we can see, the exhaustive derivation of \code{mean} only yielded a slightly smaller variance.

>>>>

The internal model case again insures the zero mean residual property over $F$, and the \textbf{point estimates} are thus calculated as

@@

Note that the \code{grisons} inventory data set originally corresponds to a classical two-phase sample design, i.e. the full set of explanatory variables is available at all sample locations in the inventory area. To demonstrate the three-phase estimation procedure, we provide an artificial three-phase scenario by recoding the phase indicators in the data set (column \code{phase_id_3p}). The values for indicating the phases (\code{0,1,2}) follow the terminology used in this article ($s_0$, $s_1$, $s_2$).

We can use the \code{table()} function to see the sample sizes of all phases for the three-phase scenario:

@@

We see that in the three-phase example, we now have 40 terrestrial sample locations providing the timber volume per hectare (\code{tvol}). The large phase $s_0$ comprises $178 + 88 + 40 = 306$ sample locations, since $s_2 \subset s_1 \subset s_0$. We now assume that at sample locations $x \in s_0$, only the mean canopy height (\code{mean}) is available, whereas at sample locations $x \in s_1$ we additionally have the explanatory variables \code{stddev}, \code{max} and \code{q75} at hand. We can now define the reduced and full regression model formulas that we will use in the \code{threephase()} function. Note that the models are nested.

@@

Note that compared to the \code{twophase()}-function, we now have to specify two regression models, i.e. the nested reduced (\code{formula.s0}) and full (\code{formula.s1}) regression model. In addition, we also have to specify the indication of the $s_1$ phase (\code{s1.id}) in the argument \code{phase_id}. The global three-phase estimation can thus be applied by

@@

The \code{summary()} of a \code{threephase()}-function recalls both regression model formulas and also gives the coefficients of determination $R^2$ for both the reduced (\code{r.squared_reduced}) and the full (\code{r.squared_full}) regression model. The set of sample sizes have also been extented by the $s_0$ phase (\code{n0}). In order to compare the results to the respective two-phase estimation, we assume that we have the mean canopy height as the only explanatory variable at all 306 sample locations:

@@

% ---------------------------------------------------------------------------- %
\subsection{Small Area Estimators}
\label{sec:threephase_sae}

@@

The \textbf{small} (\textit{small}) and \textbf{pseudo small area estimator} (\textit{psmall}) follow the same principle as in the two-phase case, i.e. applying the globally derived reduced and full regression model to the small area means of the explanatory variables, and then correcting for a potential model bias by adding the residual mean of the full model, which is not zero in general, to the point estimate (equations \ref{eq:pe_3p_small} and \ref{eq:pe_3p_psmall}). Note that the difference between the mean $\hat{\bar{\pmb{Z}}}^{(0)}_{1,G}$ derived in the $s_1$ phase and the more precise mean $\hat{\bar{\pmb{Z}}}^{(0)}_{0,G}$ or exact mean $\bar{\pmb{Z}}^{(0)}_{0,G}$ derived in the $s_0$ phase is again considered as a residual term likewise in the global estimation (equation \ref{eq:reg3p_ex} and \ref{eq:reg3p_nex}).

@@

The \textbf{synthetic} (\textit{synth}) and \textbf{pseudo synthetic estimator} (\textit{psynth}) can be applied if no terrestrial samples are available in the small area, i.e. $n_{2,G}=0$. Consequently, the residual correction and the residual variation term of the full model can no longer be applied and drops from the point (equation \ref{eq:pe_3p_synth} and \ref{eq:pe_3p_psynth}) and design-based variance (equation \ref{eq:var_3p_reg_synth} and \ref{eq:var_3p_reg_psynth}) formulas. The point estimates are again potentially biased since the zero mean residual property $\int_G R(x)dx=0$ of the full model can not be insured within the small area $G$. Also the variance will be small but to the cost of ignoring the model uncertainties. Note that there is again no external variance formula for the synthetic and pseudo synthetic estimation.

@@

The \textbf{extended synthetic} (\textit{extsynth}) and \textbf{extended pseudo synthetic estimator} (\textit{extpsynth}) insures that the residuals of the full model over both the entire inventory area $F$ and the small area $G$ are zero at the same time, i.e. $\int_F R(x)dx=\int_G R(x)dx=0$. This is again realized by extending the vector of explanatory variables by a binary categorical indicator variable $I_G(x)$ which takes the value 1 if the observation lies inside the small area $G$ and is otherwise set to 0. The extended auxiliary vector is thus defined as $\pmb{\mathbb{Z}}^t(x)=(\pmb{\mathbb{Z}}^{(0)t}(x),\pmb{\mathbb{Z}}^{t}(x))$, where $\pmb{\mathbb{Z}}^{(0)t}(x)=(\pmb{Z}^{(1)t}(x), I_G^t(x))$ and $\pmb{\mathbb{Z}}^{t}(x)=\pmb{Z}^{t}(x)$. This practically means that we have to specify the binary indicator variable for \textit{all} observations in the data frame that is passed to the package-functions.

@@

We will demonstrate the use of three-phase small area estimation in the package \pkg{forestinventory} by applying the \textbf{extended synthetic} and the \textbf{synthetic} estimator to the \code{grisons} data set. The setup is thus exactly the same as in the example for global three-phase estimation (section \ref{sec:glob_est_3p}). However, this time will use the exact auxiliary mean of the mean canopy height variable (\code{mean}) and assume that we do not know the exact means of the remaining explanatory variables \code{stddev}, \code{max} and \code{q75}. We thus first define the true means for each small area likewise we did in the \code{twophase()} example (section \ref{sec:twophase_sae}):

@@

We see that the \code{threephase()}-function returns the sample sizes in the entire inventory area as well as within each small area. The value \code{Inf} for \code{n0G} indicates that the explanatory variables at $s_0$ sample locations used in the reduced model were in our case derived exhaustively. If we compare the two results, we see that the synthetic estimation again yields a much smaller variance than the extended synthetic estimation, but to the cost of ignoring the model uncertainties.

We can also analyse how the exhaustive derivation of \code{mean} performed compared to the case where \code{mean} is non-exhaustively available but at a very large $s_0$ phase with $n_{0,G}>>n_{1,G}$. To do this, we additionally compute the \textbf{extended pseudo synthetic} estimates. As we can see, the exhaustive derivation of \code{mean} only yielded a slightly smaller variance.
